The shape of kelp: environmental influences on kelp morphology

### 2018-10-18 chapter_2.Rmd
more of Christo's additions added. some minor glitches in transferring additions have been correct. images to follow

>>>>

In Figure 2 the total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length). These values have been computed using model output from approximately 20 year of simulated nearshore wave conditions. Hindcast global wave parameters were refracted into the nearshore via a spectral phase-averaged numerical wave modelling code SWAN (Simulating, WAves in the Nearshore - 3rd generation). The model is fully spectral in all frequencies and directions (0o -360o) (can solve swell and local wind generated waves simultaneously, propagating in different direction) and was used to model the propagation of short crested swell waves, waves generated by wind, non-linear wave-wave interaction and dissipation and depth induced breaking and refraction. Wave energy dissipation due to bottom friction and whitecapping is also accounted for (Deltares, 2014).

Time series output from 1997 till 2013 of this model was then produced at the 7m and 15m depth contour lines. Each nearshore output point thus contained a 17-year timeseries of significant wave heights, peak periods and mean wave directions (with both swell and wind waves). This data set is available on a 500m resolution around most of the South African coastline and on a 200m resolution with False Bay as provided by the South African Department of Environmental Affairs (DEA) and Council for Scientific and Industrial Research (CSIR)(@heron2014). The total 17-year average energy, for each output location, was then computed using linear wave theory, taking into account deep, intermediate and shallow water wave approximations (based on wave length) for the wave energy (Holthuijsen, 2007; Kasiulis, Punys and Kofoed, 2015; Foteinis *et al.*, 2017) which is a function of wave height and period.

The directional sheltering effect of the Cape Peninsula, against the dominant swell direction, (given in Figure 1) is clear observed in the wave exposures presented in Figure 2. The classification from fully sheltered to extremely exposed is based on the total wave energy upper and lower limits. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world (Norderhaug *et al.*, 2012; Liberti, Carillo and Sannino, 2013; Sundblad *et al.*, 2014). Kelp and other ecosystems adapt to their typical environmental exposure and thus what is classified as extreme abiotic coastal forcings (temperature, winds, waves and currents) will vary around the world depending on the geographical setting.

From Figure 2 the western periphery of the Cape peninsula is almost continuously producing high coastal wave exposures while the Eastern periphery of the peninsula (western coastline of False Bay) revealed sheltered wave exposures. Here the marked seasonality, with higher energy waves during winter, may be clearly observed once more.

*Figure 2: Coastal wave energy exposure around the Cape Peninsula for the total 17-year nearshore wave record, (b) only winter months and (c) only summer months. *

In Figure 3 the propagation of a typical offshore wave spectrum is given as produced from a single time-step in SWAN. Figure 3 is presented to clarify the averaged wave exposure maps presented in Figure 2. Tracing the wave height contours into False Bay its clear why this Bay’s western periphery is predominantly sheltered. It should be mentioned that some of the annual winter frontal depression systems pass the Cape Peninsula from the west to east, resulting in wave propagating towards the continent from much more southerly directions. This results in positive and negative wave exposure anomalies all around the peninsula.

*Figure 3: SWAN numerical output for an example typical offshore wave conditions refracting around the Cape Peninsula.  *
